---
title: "Tetanus Surveillance — State of Arcadonia, 2015–2024"
author: "Arcadonia Department of Public Health"
date: today
echo: false
keep-md: true
lang: en-US
format:
  html: 
    toc: true
    axe:
      output: document
  typst:
    pdf-standard: ua-1
    margin:
      top: 1.75in
---

::: {.callout-note}
## Simulated data

All numbers, names, and findings are **simulated**. The State of
Arcadonia does not exist. Data were generated by
`data/arcadonia-tetanus.R`.
:::


```{r}
#| label: setup
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(gt)
library(readr)
library(tidyr)

cases <- read_csv("data/arcadonia-tetanus.csv", show_col_types = FALSE) |>
  mutate(
    age_group = factor(age_group, levels = c("<20", "20–64", "≥65")),
    sex = factor(sex, levels = c("F", "M"))
  )

populations <- read_csv("data/arcadonia-populations.csv", show_col_types = FALSE)
```

```{r}
#| message: false
library(brand.yml)
brand <- read_brand_yml()
base_family <- brand$typography$base$family
systemfonts::require_font(base_family)
ink <- brand$color$foreground
paper <- brand$color$background
```

## Summary

During 2015–2024, the Arcadonia Department of Public Health received
`r sum(cases$cases)` confirmed tetanus case reports — an overall
incidence of `r round(sum(cases$cases) / (2e6 * 10) * 1e5, 2)` per
100,000 person-years. Cases were concentrated among adults aged ≥65
years and among males.

## Results

### Annual case counts

Annual counts remained low throughout the period (@fig-cases), with
cases among adults aged ≥65 years exceeding those among younger
residents in most years.

```{r}
#| label: fig-cases
#| fig-cap: "Reported tetanus cases by year and age group, State of Arcadonia, 2015–2024."
#| fig-alt: "Line chart showing reported tetanus cases per year from 2015 to 2024 in Arcadonia, split into two series: people aged under 65 and people aged 65 and older. Both series stay between 1 and 4 cases per year, with a slight increase over time."
#| fig-width: 7
#| fig-height: 3.5
cases |>
  mutate(age_65 = if_else(age_group == "≥65", "≥65 years", "<65 years")) |>
  count(year, age_65, wt = cases) |>
  ggplot(aes(year, n, colour = age_65, linetype = age_65)) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_color_manual(values = c(brand$color$palette$forest, brand$color$palette$rust))+
  labs(
    x = NULL,
    y = "Reported cases",
    colour = NULL,
    linetype = NULL
  ) +
  theme_minimal(
    base_size = 12, 
    base_family = base_family, 
    ink = brand$color$foreground, 
    paper = brand$color$background
  ) +
  theme(legend.position = "top") 
```

### Incidence by age group and sex

Cases were concentrated among older adults (@tbl-demographics). Adults
aged ≥65 years had an incidence of 0.65 per 100,000 person-years —
roughly fourfold the rate in adults aged 20–64 — and males accounted
for approximately two-thirds of reported cases.

```{r}
#| label: tbl-demographics
#| tbl-cap: "Reported tetanus cases and incidence rates by age group and by sex — State of Arcadonia, 2015–2024."

cases |>
  pivot_longer(c(age_group, sex), names_to = "section",
               values_to = "level", values_transform = as.character) |>
  count(section, level, wt = cases, name = "n_cases") |>
  left_join(populations, by = "level") |>
  mutate(
    section = recode(section, age_group = "Age", sex = "Sex"),
    rate_per_100k = round(n_cases / (population * 10) * 1e5, 2)
  ) |>
  select(section, level, population, n_cases, rate_per_100k) |>
  gt(groupname_col = "section", rowname_col = "level") |>
  cols_label(
    population = "Population",
    n_cases = "Cases",
    rate_per_100k = "Rate"
  ) |>
  tab_footnote(
    footnote = "per 100,000 person-years",
    locations = cells_column_labels(columns = rate_per_100k)
  ) |> 
  fmt_number(columns = population, decimals = 0, use_seps = TRUE) |>
  cols_align(align = "right", columns = -level) |>
  theme_brand_gt() |> 
  opt_table_font(base_family, weight = "regular", add = FALSE) |> 
  tab_options(
    heading.title.font.weight    = "normal",
    heading.subtitle.font.weight = "normal",
    stub.font.weight             = "normal",
    stub_row_group.font.weight   = "normal",
    row_group.font.weight        = "normal"
  )
```

## Background

Tetanus is caused by toxin from *Clostridium tetani* contaminating
wounds. Disease is rare where childhood vaccination coverage is high,
but adults with waning immunity — especially those aged ≥65 years —
remain at risk. In Arcadonia, adult decennial booster coverage has
plateaued near 60%.

## Methods

Confirmed tetanus cases (acute illness with muscle spasms or
hypertonia, without another apparent cause, reported by a licensed
clinician) reported via passive clinician reporting during 2015–2024
were tabulated by year, age group, and sex. Age- and sex-specific
incidence rates were computed using estimated Arcadonia population
denominators.

## Discussion

The Arcadonia pattern mirrors national surveillance: cases rare,
older-skewed, male-predominant. Continued emphasis on adult booster
vaccination — particularly in adults aged ≥65 years — and prompt wound
management remains warranted. Passive reporting may underrepresent
mild cases; population denominators are estimates.


